---
title: 'Rest Activity Rhythms During COVID19 Pandemic'
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: inline
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
.main-container {
    max-width: 1800px !important;
  }
</style>

```{r global_options, include=FALSE}
#options(width = 1300)
knitr::opts_chunk$set(warning=FALSE)
```

Author: TeYang, Lau <br>
Last Updated: 6 June 2020

<img src = './Pictures/covid.jpg'>

# 1. Data Loading
```{r, message=F,warning=F}
## Load Packages 
library(tidyverse); library(knitr); library(kableExtra); library(formattable); library(patchwork); library(zoo);
library(GGally); library(ggExtra); library(stargazer); library(car); library(factoextra); library(NbClust);
library(ggplot2); library(scales); library(ggforce); library(ggpubr); library(gmodels); library(vcd); 
library(rstatix); library(ez); library(dendextend); library(hrbrthemes); library(emmeans); library(psych)


## Load Data
steps <- read.csv('./Data/steps_wide_cleaned_JANtoAPR2020_allsubjs_validdays.csv', header = TRUE)
L2 <- read.csv('./Data/Euclideank4k5.csv', header = F) %>%
  rename('cluster_k4_L2' = 'V1', 'cluster_k5_L2' = 'V2') %>% select(cluster_k4_L2)

# Reorder cluster assignments so that clusters are in ascending order in N size
L2$cluster_k4_L2 <- match(L2$cluster_k4_L2, names(sort(table(L2$cluster_k4_L2))))
L2$cluster_k4_L2 <- match(L2$cluster_k4_L2, names(sort(table(L2$cluster_k4_L2))))
L2$cluster_k4_L2 <- match(L2$cluster_k4_L2, c(1,3,2,4)) # swap 3 and 2 so that 2 is middle and 3 is later group

## Rename columns
tod_15min <- as.character(read.table("./Data/tod_15min.txt")[,1]) # load text file with 24hr time in 15min intervals
names(steps)[3:98] <-c(tod_15min)

head(steps[,1:24])
```
<br>

# 2. Data Cleaning
<font size="4"> 

1. Steps Data on Weekdays
  
2. First 28 days
  
3. At least 10 Weekdays
  
4. Filter days out: 
  
    + Total Steps > 50000 
    + Total Steps > 40000 & Sedentary Minutes > 1320 min
    + Sedentary Minutes == 1440 min
    + No Resting Heart Rate
    + WearTime < 780 min

</font>

# 3. Kmeans Clustering Steps
<br>
```{r, message=F,warning=F}
## Log Steps
# log steps and replace -Inf with 0
steps_log <- log(steps[-1:-2]) %>% anchors::replace.value(., c(1:96), from=-Inf, to=0) %>% 
  mutate(DRID = steps$DRID, Date = steps$Date) %>% select(DRID,Date,everything())

steps_log <- cbind(steps_log, L2) %>% mutate(cluster_k4_L2=factor(cluster_k4_L2))

```
<br>

## 3.1. Rest Activity Rhythm Profile Plots
```{r, message=F,warning=F}

# Euclidean cluster aggregate
steps_agg_log_k4_L2 <- aggregate(steps_log[, 3:98],
                           by = list(steps_log$cluster_k4_L2),
                           FUN = mean)



## Plot Cluster Profiles
#### Euclidean Distance 

color4 <- c("chartreuse2","dodgerblue2",'tan3',"#990000")

par(mar=c(5,6,3,2))
plot(0,0,xlim = c(0,95),ylim = c(0,5.5),type = "n", xlab='Time of Day', ylab='Log Steps/15 min', xaxt="n", yaxt='n', cex.lab=2) # create empty plot for line graphs
axis(1, at=c(0,47,95),labels=c('12am', '12pm', '12am'),cex.axis=2, tck=-0.02)
for (i in 1:4){
  lines(c(0:95),steps_agg_log_k4_L2[i,-1], col = color4[steps_agg_log_k4_L2$Group.1[i]], lwd = 3)
}
title(main="K-means Euclidean Distance Clustering\nK=4")
legend("topleft", legend=c(paste('Active 3-Peak Early, days =',as.character(table(steps_log$cluster_k4_L2)[1]),sep=' '),
                           paste('3-Peak Middle, days =',as.character(table(steps_log$cluster_k4_L2)[2]),sep=' ')),
       col=color4, lty=1, lwd=3, cex=1, bty='n')
legend("bottomright", legend=c(paste('Active 2-Peak Later, days =',as.character(table(steps_log$cluster_k4_L2)[3]),sep=' '),
                           paste('Inactive 3-Peak, days =',as.character(table(steps_log$cluster_k4_L2)[4]),sep=' ')),
       col=color4[3:4], lty=1, lwd=3, cex=1, bty='n')
axis(2, cex.axis=2, tck=-0.02)

```

## 3.2. Individual Time Series RAR Plots
```{r, message=F,warning=F}
# Get some samples for plotting
samp <- c("439CD3EEBF37", "13F9D36BC9AC", "63A98BD38B70")
# "CCC8BC39C073" "B7891B5C3B02" "D0C79BF6A82A" "439CD3EEBF37" "13F9D36BC9AC" "63A98BD38B70"
sampledf <-  data.frame()
for(i in 1:length(samp)) {
  id <- samp[i]
  subjnum <- which(as.character(unique(steps_log$DRID)) %in% id)
  indiv_steps <- filter(steps_log, DRID==id) %>% arrange(Date)
  indiv_steps_long <- gather(indiv_steps, Time, Steps, 3:98) %>% arrange(Date,Time) %>% mutate(DateTime = as.POSIXct(paste(Date,Time)))
  indiv_steps_long <- indiv_steps_long %>% mutate(index=rep(subjnum, dim(indiv_steps_long)[1]))
  sampledf <- rbind(sampledf, indiv_steps_long)
}

# Plot
par(mar=c(5,6,3,6))
p <- ggplot(sampledf, aes(DateTime, Steps, color=cluster_k4_L2)) + geom_path(aes(group=1)) +
          theme_classic() +
          ylab('Log Steps/15 min') +
          #scale_fill_manual(name = "Cluster", breaks = c("Delayed", "Most Active", "Least Active")) +
          scale_colour_manual('Profiles',values=color4, breaks=c("1", "2", "3", "4"), labels=c("Active 3-Peak Early", "3-Peak Middle", "Active 2-Peak Later", "Inactive 3-Peak"),
        guide=guide_legend(keywidth=3, override.aes = list(size = 2)),                              ) +
          scale_x_datetime(labels=date_format('%d %b', tz='Singapore'), date_breaks = '7 days', expand=c(0,0.9)) +
          theme(axis.text.x=element_text(angle=45, hjust=1, size=10), legend.key.size = unit(1, "cm"),legend.title = element_text('Profiles', size=15, face="bold"), 
        legend.text = element_text(size=15), axis.text.y = element_text(size=15), axis.title=element_text(size=15), 
        plot.title = element_text(hjust = 0.5)) + 
  geom_vline(xintercept=as.numeric(as.POSIXct(c('2020-03-26','2020-04-7'))), linetype='dashed', lwd=1, color = "black")

p + facet_grid(rows=vars(index))
  
```
<br>

# 4. Hierarchical Clustering of Proportion of Time Spent in each Profile

## 4.1. Clean and wrangle
```{r, message=F,warning=F}
# Fill in missing days
uniqueID = unique(steps_log$DRID)
DRID = rep(uniqueID, each = 118)
Date = rep(seq(as.Date('2020-01-01'),as.Date('2020-04-27'), by='1 day'), times = length(unique(steps_log$DRID)))
fulldays = data.frame(DRID, Date) # create dataframe with full 118 days since day start for each ID

steps_period <- steps_log %>% select(DRID,Date,cluster_k4_L2) %>% 
                           mutate(Date = as.Date(as.character(Date))) %>% filter(Date <= '2020-04-27')
alldaysteps <- full_join(fulldays, steps_period)

# Filter Ps with enough days
preeligible <- steps_log %>% mutate(Date=as.Date(as.character(Date))) %>% filter(Date < '2020-04-07') %>% group_by(DRID) %>% summarise(predays = n()) %>% 
  filter(predays >= 59) %>% as.data.frame() # 59/97 days for pre-cb 60%
posteligible <- steps_log %>% mutate(Date=as.Date(as.character(Date))) %>% filter(Date >= '2020-04-07' & Date <= '2020-04-27') %>%
  group_by(DRID) %>% summarise(postdays = n()) %>% filter(postdays >= 13) %>% as.data.frame() # 13/21 days for pre-cb
eligible <- merge(preeligible, posteligible) 
#eligible <- filter(eligible, !DRID %in% c('A4596E5A98F6', 'F2D312B33D63'))

alldaysteps <- filter(alldaysteps, DRID %in% eligible$DRID)

long <- alldaysteps %>% mutate(Date = as.POSIXct(Date, format='%Y-%m-%d', tz='Singapore'))

events <- data.frame(n=c('First\nCase','Chinese\nNew Year','Health Alert\nLevel Raised','WHO Pandemic','Malaysia\nBorder Closure','Pubs Closure*', 'Circuit Breaker'), Date=as.POSIXct(c('2020-01-21','2020-01-27','2020-02-08','2020-03-08','2020-03-17','2020-03-26','2020-04-7')))

```
<br>

## 4.2. Proportion of time spent in each profile
```{r, message=F,warning=F}
# K4_L1

# Proportion of time spent in each state Pre-Circuit Breaker
precb <- long %>% filter(Date <'2020-04-07') %>% filter(!is.na(cluster_k4_L2)) %>% group_by(DRID) %>% 
  mutate(ndays=n()) %>% as.data.frame() %>% # get total days for each subject with no NAs
  group_by(DRID,cluster_k4_L2) %>% 
  summarize(proportion=n()/ndays[1]) %>% as.data.frame() %>%
  spread(., cluster_k4_L2, proportion) 
precb[is.na(precb)] <- 0
names(precb)[2:5] <-c("Pre-CB Active 3-Peak Early", "Pre-CB 3-Peak Middle",'Pre-CB Active 2-Peak Later', "Pre-CB Inactive 3-Peak")

# Proportion of time spent in each profile During Circuit Breaker
duringcb <- long %>% filter(Date >= '2020-04-07') %>% filter(!is.na(cluster_k4_L2)) %>% group_by(DRID) %>% 
  mutate(ndays=n()) %>% as.data.frame() %>% # get total days for each subject with no NAs
  group_by(DRID,cluster_k4_L2) %>% 
  summarize(proportion=n()/ndays[1]) %>% as.data.frame() %>%
  spread(., cluster_k4_L2, proportion)
duringcb[is.na(duringcb)] <- 0
names(duringcb)[2:5] <-c("CB Active 3-Peak Early",  "CB 3-Peak Middle", 'CB Active 2-Peak Later', "CB Inactive 3-Peak")

preduringcb_k4_L2 <- merge(precb,duringcb,by='DRID')
preduringcb_k4_L2 <- preduringcb_k4_L2[,c(1,2,6,3,7,4,8,5,9)]

#z-score features
# preduringcb_k4_L2 <- preduringcb_k4_L2 %>% select(-DRID) %>% scale(.) %>% as.data.frame() %>% 
#   mutate(DRID = precb$DRID) %>% select(DRID, everything())

head(preduringcb_k4_L2)
```